---
author:
output:
  html_document: default
  pdf_document: default
  word_document: default
bibliography: references.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)
```

Date:

```{r, echo = FALSE}
d <- read.csv("chronic_pain.csv", header = TRUE, sep = ";")

## I can put in here all analyses I would like to refer to later.
# fit <- lm()
```

---
title: "Association of socio-demographic factors and pain presentation in a cohort of chronic pain patients"
---

## Introduction

Chronic pain is a complex and multifaceted condition that manifests in highly heterogeneous ways across individuals. Much variability is observed in the intensity of pain, the response to conservative treatment, and scores on other related patient-reported outcome measures (PROMs) such as anxiety and depression. The aim of this study was to estimate associations between socio-demographic characteristics and pain intensity in patients with chronic pain at the intake in a specialized pain clinic. The secondary aim was to estimate correlations between PROMs in patients with chronic pain.

## Methods
For this study, we used a subset of patients from the chronic pain registry of the Maastricht University Medical Centre department of pain management who had been recruited between January 2022 and December 2024. This registry has been described in great detail elsewhere. Patients were considered eligible if they had pain for at least 3 months and conservative treatment (e.g. physical therapy) failed. Patients received questionnaires at enrollment, consisting of several PROMs, such as the RAND-36 questionnaire and the hospital anxiety and depression scale (HADS). Baseline data were summarized as mean and standard deviation (SD) or as count and percentage. Multivariable linear regression analysis was used to estimate associations between socio-demographic factors and pain intensity. Pearson's correlation coefficient was computed to estimate the correlation between pain intensity, depressive complaints, and anxiety.

## Results
### Baseline characteristics
Between January 2022 and December 2024, we included chronic pain patients, of whom (%) were female, with a mean age (SD) of (). Table 1 shows baseline characteristics of the cohort of chronic pain patients.

### Associations between socia-demographic factors and pain intensity
The mean (SD) pain intensity measured using the RAND questionnaire pain domain was (). Linear regression analysis did not reveal any statistically significant associations between socio-demographic factors and pain intensity (table 2). However, both female sex (regression coefficient = ) and having a child (regression coefficient = ) were associated with a clinically meaningful higher mean pain intensity, albeit not significant.

```{r, echo = FALSE, message = FALSE, results = 'asis'}
library(stargazer)

# Make sure to pick the piece of code that fits the way you knit the report (pdf or HTML)
# Note that the R code chunk header includes results = 'asis'. Find out what this does!

# stargazer()
```

### Correlations between outcome measures
Figure 1 shows the correlations between different outcome measures. Strong negative correlations were seen between the HADS depression and anxiety scales on the one hand, and the RAND questionnaire mental health dimension, and moderate positive correlations between the two HADS subscales. The RAND pain dimension was not strongly correlated to any of the other patient-reported outcomes.

```{r, echo = FALSE, message = FALSE, fig.cap=""}

library(corrplot)

# corr <- cor(d[, 6:17], use = "pairwise.complete.obs")
# colnames(corr) <- c()
# rownames(corr) <- colnames(corr)
# corrplot(corr)

```

## Conclusion
We did not find strong significant associations between socio-demographic factors and pain intensity in a large cohort of chronic pain patients, albeit some were potentially clinically relevant. Pain intensity was not strongly correlated to other patient outcomes.

## References
