---
title: "COVID time series by HDI(Europe)"
author: "Lin"
date: "2026-02-26"
output:
  html_document: default
  pdf_document: default
---

## This analysis uses COVID data from European nations.
## Countries are classified into three approximately tertile-based HDI groups(Low, Mid, and High HDI) using country-level Human Development Index (HDI) values,while keeping identical HDI values in the same group.
## For each country, the analysis uses total_cases_per_million at the end of each month, and then calculates monthly group means and plots a time-series figure with 95% CI.

```{r setup echo = FALSE, message = FALSE, warning = FALSE}
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)

## Read the cleaned COVID dataset
PATH_COVID <- file.path("open_covid_data", "Covid_cleaned.csv")
covid_raw <- readr::read_csv(PATH_COVID, show_col_types = FALSE)
```

```{r rename-mutate echo = FALSE, message = FALSE, warning = FALSE}
## Rename variables and standardise data types

recodedCovid <- covid_raw %>%
  dplyr::rename(country = location, num_total_cases_per_million = total_cases_per_million, num_hdi = human_development_index
                ) %>%
dplyr::mutate(date = as.Date(date), continent = as.character(continent), country = as.character(country), num_total_cases_per_million = as.numeric(num_total_cases_per_million), num_hdi = as.numeric(num_hdi))
```

```{r filter-Europe echo = FALSE, message = FALSE, warning = FALSE}
## keep European countries only and remove missing value

recodedCovid <- recodedCovid %>%
  dplyr::filter(continent == "Europe", !is.na(date), !is.na(country), !is.na(num_total_cases_per_million), !is.na(num_hdi))
```

```{r make-month echo = FALSE, message = FALSE, warning = FALSE}
## create a monthly time variable 

recodedCovid <- recodedCovid %>%
  dplyr::mutate(month = lubridate::floor_date(date,"month"))
```

```{r create-hdi-groups echo = FALSE, message = FALSE, warning = FALSE}
## Compute country-level mean HDI
hdiCountry <- recodedCovid %>%
  dplyr::group_by(country) %>%
  dplyr::summarise(num_hdi_country = mean(num_hdi, na.rm = TRUE),.groups = "drop")

## Build a frequency table of unique HDI values so tied values stay together 
hdi_freq <- hdiCountry %>%
  dplyr::count(num_hdi_country, name = "freq") %>%
  dplyr::arrange(num_hdi_country) %>%
  dplyr::mutate(cum_n = cumsum(freq))

## Total number of countries and target tertile positions
N <- sum(hdi_freq$freq)
target1 <- N / 3 
target2 <- 2 * N / 3

## Pick the first cut at the boundary closest to the 1/3 position
i1 <- which.min(abs(hdi_freq$cum_n - target1))

## Pick the second cut at the boundary closest to the 2/3 position
candidate_i2 <- seq(i1 + 1, nrow(hdi_freq) - 1)
i2 <- candidate_i2 [which.min(abs(hdi_freq$cum_n[candidate_i2] - target2))]

## Extract cut values (upper boundary for Low and Mid groups)
cut1 <- hdi_freq$num_hdi_country[i1]
cut2 <- hdi_freq$num_hdi_country[i2]

## Assign groups without splitting identical HDI values
hdiCountry <- hdiCountry %>%
  dplyr::mutate(hdi_group = dplyr::case_when(
    num_hdi_country <= cut1 ~ "Low HDI",
    num_hdi_country <= cut2 ~ "Mid HDI", TRUE ~ "High HDI"), hdi_group = factor(hdi_group, levels = c("Low HDI", "Mid HDI", "High HDI")))

## join country-level HDI group lables back to daily data
recodedCovid <- recodedCovid %>%
  dplyr::left_join ( hdiCountry %>%
  dplyr::select(country,hdi_group), by = "country")
```


```{r month-end-country echo=FALSE, message=FALSE, warning=FALSE}
## For each country and month, keep the last available total_cases_per_million value
countryMonth <- recodedCovid %>%
  dplyr::group_by(country, hdi_group, month) %>%
  dplyr::arrange(date, .by_group = TRUE) %>%
  dplyr::summarise(
    month_end_cases_pm = dplyr::last(num_total_cases_per_million), 
.groups = "drop")
```


```{r group-month-summary echo=FALSE, message=FALSE, warning=FALSE}
## Summarise month-end values by HDI group and month:mean, standard deviation, standard error, and 95% confidence interval

groupMonth <- countryMonth %>%
  dplyr::group_by(hdi_group, month) %>%
  dplyr::summarise(
    mean_value = mean(month_end_cases_pm, na.rm = TRUE),
    sd_value   = sd(month_end_cases_pm, na.rm = TRUE),
    n_countries = dplyr::n(),
    se = sd_value / sqrt(n_countries), 
    lower_95 = mean_value - 1.96 * se, 
    upper_95 = mean_value + 1.96 * se, 
    .groups = "drop"
  ) %>%
  dplyr::arrange(hdi_group, month)
```


```{r final-plot echo=FALSE, message=FALSE, warning=FALSE}
## Plot monthly mean total_cases_per_million by HDI group, using ggplot2, with lines for means and shaded for 95% CI

p_final <- ggplot(
  groupMonth, 
  aes(x = month, y = mean_value, colour = hdi_group, fill = hdi_group)) + 
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95), 
              alpha = 0.18, colour = NA, na.rm = TRUE) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  scale_x_date(date_labels = "%b %Y", date_breaks = "1 month") +
  labs(x = "Month", y = "Mean total_cases_per_million (month-end, by country)", title = "Europe: total_cases_per_million by HDI group", colour = "HDI group", fill = "HDI group") +
  theme_bw() + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

dir.create(file.path("outputs","figures"), recursive = TRUE, showWarnings = FALSE)
ggsave(file.path("outputs","figures","hdi_timeseries.png"), p_final, width = 9, height = 5, dpi = 300)

p_final
```
