---
title: "Final Group Assignment"
subtitle: "EPI4934B Clinical Data Science"
author: "Paulina Stapor (i6325887), Shiqiu Meng (i6251461), Lin Yuan (i6420954)"
date: "6 March 2026"
output:
  html_document:
    theme: flatly
    toc: true
    toc_float: true
    number_sections: true
---

<style>
body {
  margin-bottom: 0 !important;
  padding-bottom: 0 !important;
}

.main-container {
  margin-bottom: 0 !important;
  padding-bottom: 0 !important;
}

p:last-child, div:last-child {
  margin-bottom: 0 !important;
  padding-bottom: 0 !important;
}
</style>

-----

<div style="border:4px solid #2e7d32; padding:20px; border-radius:10px; background:#e8f5e9; font-size:16px;">

🟢🟢🟢 **IMPORTANT REPRODUCIBILITY NOTICE — PLEASE READ BEFORE KNITTING** 🟢🟢🟢

To ensure that this report can be reproduced exactly, the analysis depends on the **specific R package environment recorded in `renv.lock`.**

Before knitting or rendering this document, please **restore the project environment** by running:

```r
renv::restore()
```

This step installs the **exact package versions used during development**.

If the environment is not restored first, the report may:

• fail to render  
• encounter missing package errors  
• produce results or figures that differ from the original analysis  

Thank you for restoring the environment before running the report.

</div>


```{r pdf-widget-setup, include=FALSE}
# Make knitr convert HTML widgets to images when knitting to PDF
knitr::opts_chunk$set(dev = "png")

# Ensure all file paths are relative to the project root
knitr::opts_knit$set(root.dir = here::here())

if (knitr::is_latex_output()) {
  # Helps htmlwidgets behave in non-HTML output
  options(htmlwidgets.TOHTML_ARGS = list(standalone = FALSE))
}
```


```{r setup-renv, include=FALSE}
# ------------------------------------------------------------------
# renv dependency hint
#
# `rnaturalearthdata` is used indirectly by
# rnaturalearth::ne_countries(). Because packages in this project
# are loaded dynamically via a loop over `required_packages`,
# renv::dependencies() cannot detect this dependency automatically.
#
# Referencing the package namespace below allows renv to detect it
# when renv::snapshot() scans the project.
# ------------------------------------------------------------------

rnaturalearthdata::countries110
```






# Global Trends in monthly New COVID-19 Cases per Million by HDI category by Lin

Global trends in monthly new COVID-19 cases per million population by Human Development Index (HDI) category from October 2020 to October 2021. Countries are grouped according to the UNDP HDI classification (low, medium, high, and very high). Lines represent the monthly mean number of reported new COVID-19 cases per million population within each HDI category, and shaded areas indicate 95% confidence intervals.

```{r hdi-analysis, echo=TRUE, message=FALSE, warning=FALSE}

# Load required libraries
library(readr)
library(dplyr)
library(lubridate)
library(ggplot2)

# 1) Read cleaned COVID dataset
PATH_COVID <- file.path("open_covid_data", "Covid_cleaned.csv")
covid_raw <- readr::read_csv(PATH_COVID, show_col_types = FALSE)

# 2) Rename variables and set types
recodedCovid <- covid_raw %>%
  dplyr::rename(
    country = location,
    num_new_cases_per_million = new_cases_per_million,
    num_hdi = human_development_index
  ) %>%
  dplyr::mutate(
    date = as.Date(date),
    continent = as.character(continent),
    country = as.character(country),
    num_new_cases_per_million = as.numeric(num_new_cases_per_million),
    num_hdi = as.numeric(num_hdi)
  )

# 3) Keep country rows (all continents) and required values
recodedCovid <- recodedCovid %>%
  dplyr::filter(
    !is.na(continent),
    !is.na(date),
    !is.na(country),
    !is.na(num_new_cases_per_million),
    !is.na(num_hdi)
  )

# 4) Create monthly time variable
recodedCovid <- recodedCovid %>%
  dplyr::mutate(month = lubridate::floor_date(date, "month"))

# 5) Compute one HDI value per country (mean across available rows) + HDI groups
hdiCountry <- recodedCovid %>%
  dplyr::group_by(country) %>%
  dplyr::summarise(
    num_hdi_country = mean(num_hdi, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    hdi_group = dplyr::case_when(
      num_hdi_country < 0.550 ~ "Low HDI",
      num_hdi_country < 0.700 ~ "Medium HDI",
      num_hdi_country < 0.800 ~ "High HDI",
      num_hdi_country >= 0.800 ~ "Very high HDI",
      TRUE ~ NA_character_
    ),
    hdi_group = factor(
      hdi_group,
      levels = c("Low HDI", "Medium HDI", "High HDI", "Very high HDI")
    )
  )

# Join HDI group back to daily data
recodedCovid <- recodedCovid %>%
  dplyr::left_join(
    hdiCountry %>% dplyr::select(country, hdi_group),
    by = "country"
  ) %>%
  dplyr::filter(!is.na(hdi_group))

# 6) Within each country, compute monthly mean of daily new cases per million
countryMonth <- recodedCovid %>%
  dplyr::group_by(country, hdi_group, month) %>%
  dplyr::summarise(
    monthly_mean_new_cases_pm = mean(num_new_cases_per_million, na.rm = TRUE),
    .groups = "drop"
  )

# 7) Within each HDI group and month, compute unweighted country average + 95% CI
groupMonth <- countryMonth %>%
  dplyr::group_by(hdi_group, month) %>%
  dplyr::summarise(
    mean_value = mean(monthly_mean_new_cases_pm, na.rm = TRUE),
    sd_value = sd(monthly_mean_new_cases_pm, na.rm = TRUE),
    n_countries = dplyr::n(),
    se = sd_value / sqrt(n_countries),
    lower_95 = mean_value - 1.96 * se,
    upper_95 = mean_value + 1.96 * se,
    .groups = "drop"
  ) %>%
  dplyr::arrange(hdi_group, month)

# 8) Plot global monthly trend by HDI category
p_final <- ggplot(groupMonth, aes(x = month, y = mean_value, colour = hdi_group, fill = hdi_group)) +
  geom_ribbon(aes(ymin = lower_95, ymax = upper_95),
              alpha = 0.18, colour = NA, na.rm = TRUE) +
  geom_line(linewidth = 1.1, na.rm = TRUE) +
  scale_x_date(
    date_breaks = "1 months",
    labels = function(x) paste0(month.abb[lubridate::month(x)], ". ", lubridate::year(x))
  ) +
  labs(
    x = "Month",
    y = "mean new cases per million",
    title = "Global Trends in monthly New COVID-19 Cases per Million by HDI category",
    colour = "HDI category",
    fill = "HDI category"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold")
  )

# Create output folder
dir.create("Outputs_Time_Series_HDI_by_Lin", recursive = TRUE, showWarnings = FALSE)

# Save figure
ggsave(
  filename = file.path("Outputs_Time_Series_HDI_by_Lin", "global_hdi_newcases_monthlymean_unweighted.png"),
  plot = p_final,
  width = 16, height = 6, dpi = 300
)

# Print plot
p_final

```



# COVID-19 New Cases per Million (7-day moving average) by Paulina

Trends in new COVID-19 cases per million population (7-day moving average) by continent from 1 October 2020 to 1 October 2021. Countries are pooled at the continent level, and estimates are population-weighted.

```{r continent-analysis, echo=TRUE, message=FALSE, warning=FALSE}

# Load required libraries
library(readr)
library(dplyr)
library(ggplot2)
library(slider)
library(scales)

# Load dataset
Covid_cleaned <- read_csv(here::here("open_covid_data", "Covid_cleaned.csv"))

# Compute continent-level daily cases
continent_daily <- Covid_cleaned %>%
  mutate(
    est_new_cases = (new_cases_per_million / 1e6) * population
  ) %>%
  group_by(continent, date) %>%
  summarise(
    new_cases = sum(est_new_cases, na.rm = TRUE),
    pop = sum(population, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  mutate(
    new_cases_per_million_cont = (new_cases / pop) * 1e6
  ) %>%
  group_by(continent) %>%
  arrange(continent, date) %>%
  mutate(
    new_cases_pm_7dma = slider::slide_dbl(
      new_cases_per_million_cont,
      mean,
      .before = 6,
      .complete = TRUE,
      na.rm = TRUE
    )
  ) %>%
  ungroup()

# Remove rows with NA moving averages
continent_daily <- continent_daily %>%
  filter(!is.na(new_cases_pm_7dma))

# Plot results
ggplot(continent_daily, aes(x = date, y = new_cases_pm_7dma, color = continent)) +
  geom_line(linewidth = 1.1, alpha = 0.95) +
  scale_x_date(date_breaks = "6 months", date_labels = "%b\n%Y") +
  scale_y_continuous(labels = label_number(accuracy = 1)) +
  labs(
    title = "COVID-19 New Cases per Million (7-day moving average)",
    subtitle = "Countries pooled to continent level (population-weighted)",
    x = "Date",
    y = "New cases per million (7DMA)",
    color = "Continent"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    plot.title = element_text(face = "bold", size = 14),
    plot.subtitle = element_text(size = 13),
    legend.position = "bottom",
    panel.grid.minor = element_blank()
  )

# Create output folder
dir.create("Outputs_Time_Series_Continent_by_Paulina", recursive = TRUE, showWarnings = FALSE)

# Save the last plotted ggplot
ggsave(
  filename = file.path(
    "Outputs_Time_Series_Continent_by_Paulina",
    "continent_newcases_per_million_7dma.png"
  ),
  plot = ggplot2::last_plot(),
  width = 12,
  height = 6,
  dpi = 300
)
```



# Cumulative COVID-19 deaths per million by Shiqiu (Q)

Global distribution of cumulative COVID-19 deaths per million population as of 1 October 2021. Countries are coloured according to the cumulative number of reported COVID-19 deaths per million inhabitants, with darker colours indicating higher mortality. Mortality rates are standardised per million population to enable comparison across countries with different population sizes. Data represent the most recent available values up to 1 October 2021. Countries shown in grey indicate locations for which no data were available or were not reported in the dataset at the time of analysis.

```{r cumulative-deaths-choropleth-map, echo=TRUE, message=FALSE, warning=FALSE, out.width="100%", fig.align="left"}
# =========================================================
# Cumulative COVID-19 Deaths per Million, Interactive Maps
#   - Flat world + Rotatable globe
#   - Uses total_deaths_per_million at each country's latest date
#
# Path principles (no hard-coded script folder name):
#   - Input read from PROJECT ROOT via here::here(...)
#   - Output written next to this script via knitr::current_input()
# =========================================================


# =========================================================
# 0) Packages (install if missing; load quietly)
# =========================================================
required_packages <- c(
  "readr", "dplyr", "here", "countrycode",
  "plotly", "rnaturalearth", "rnaturalearthdata", "sf",
  "htmlwidgets", "knitr", "rstudioapi",
  "webshot2"
)

for (pkg in required_packages) {
  if (!requireNamespace(pkg, quietly = TRUE)) {
    stop(
      "Missing package: ", pkg, "\n",
      "Run renv::restore() (or install.packages('", pkg, "') ) before knitting."
    )
  }
  suppressPackageStartupMessages(library(pkg, character.only = TRUE))
}


# =========================================================
# 1) Output path helpers (no hard-coding)
# =========================================================
get_this_script_dir <- function() {
  p <- knitr::current_input()
  if (!is.null(p) && nzchar(p)) {
    return(normalizePath(dirname(p), winslash = "/", mustWork = FALSE))
  }
  
  if (rstudioapi::isAvailable()) {
    ctx <- rstudioapi::getActiveDocumentContext()
    if (!is.null(ctx$path) && nzchar(ctx$path)) {
      return(normalizePath(dirname(ctx$path), winslash = "/", mustWork = FALSE))
    }
  }
  
  normalizePath(getwd(), winslash = "/", mustWork = FALSE)
}

out_dir <- file.path(
  get_this_script_dir(),
  "Outputs_Choropleth_Map_by_Shiqiu"
)

out_path <- function(...) {
  file.path(out_dir, ...)
}

if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)


# =========================================================
# 2) Small helpers (robustness)
# =========================================================
first_non_na <- function(x) {
  x2 <- x[!is.na(x)]
  if (length(x2) == 0) NA else x2[[1]]
}


# =========================================================
# 3) Read dataset (PROJECT ROOT)
# =========================================================
csv_path <- here::here("open_covid_data", "Covid_cleaned.csv")
if (!file.exists(csv_path)) stop("File not found: ", csv_path)

df <- readr::read_csv(csv_path, show_col_types = FALSE)

needed_cols <- c("location", "continent", "date", "total_deaths_per_million")
missing_cols <- setdiff(needed_cols, names(df))
if (length(missing_cols) > 0) {
  stop("Missing required columns in CSV: ", paste(missing_cols, collapse = ", "))
}

df <- df %>%
  dplyr::mutate(date = as.Date(date))


# =========================================================
# 4) Name -> ISO3 conversion (with minimal targeted fixes)
# =========================================================
name_fixes <- c(
  "Timor" = "Timor-Leste",
  "Congo" = "Republic of the Congo",
  "Congo (Brazzaville)" = "Republic of the Congo",
  "Congo (Kinshasa)" = "Democratic Republic of the Congo",
  "DR Congo" = "Democratic Republic of the Congo",
  "Czechia" = "Czech Republic",
  "Cape Verde" = "Cabo Verde",
  "Swaziland" = "Eswatini",
  "Ivory Coast" = "Côte d’Ivoire",
  "Cote d'Ivoire" = "Côte d’Ivoire",
  "Cote dIvoire" = "Côte d’Ivoire",
  "Myanmar (Burma)" = "Myanmar",
  "North Macedonia" = "Macedonia"
)

df <- df %>%
  dplyr::mutate(
    location_clean = dplyr::if_else(
      location %in% names(name_fixes),
      unname(name_fixes[location]),
      location
    ),
    iso_code = countrycode::countrycode(
      sourcevar   = location_clean,
      origin      = "country.name",
      destination = "iso3c",
      warn        = TRUE
    )
  )

stopifnot("iso_code" %in% names(df))


# =========================================================
# 5) Location-level diagnostics (why rows become unmapped)
# =========================================================
loc_diag <- df %>%
  dplyr::group_by(location, location_clean) %>%
  dplyr::summarise(
    continent_any   = any(!is.na(continent)),
    iso_code_any    = first_non_na(iso_code),
    has_iso         = any(!is.na(iso_code)),
    has_deaths_pm   = any(!is.na(total_deaths_per_million)),
    deaths_pm_any   = first_non_na(total_deaths_per_million),
    latest_date_any = suppressWarnings(max(date, na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  dplyr::mutate(
    reason = dplyr::case_when(
      !continent_any             ~ "Excluded by filter: continent is NA (aggregate/non-country row)",
      !has_iso                   ~ "ISO conversion failed (countrycode could not map name to ISO3)",
      has_iso & !has_deaths_pm   ~ "total_deaths_per_million missing in dataset for this country",
      TRUE                       ~ "OK"
    )
  )

loc_diag_iso <- loc_diag %>%
  dplyr::filter(!is.na(iso_code_any)) %>%
  dplyr::group_by(iso_code_any) %>%
  dplyr::summarise(
    any_deaths_missing        = any(reason == "total_deaths_per_million missing in dataset for this country", na.rm = TRUE),
    any_excluded_by_continent = any(reason == "Excluded by filter: continent is NA (aggregate/non-country row)", na.rm = TRUE),
    any_iso_failed            = any(reason == "ISO conversion failed (countrycode could not map name to ISO3)", na.rm = TRUE),
    .groups = "drop"
  )


# =========================================================
# 6) World polygons index (Natural Earth)
#   - Uses iso_a3 with adm0_a3 fallback
# =========================================================
world_sf <- rnaturalearth::ne_countries(scale = "medium", returnclass = "sf")

world_index <- world_sf %>%
  sf::st_drop_geometry() %>%
  dplyr::transmute(
    iso_a3 = dplyr::if_else(is.na(iso_a3) | iso_a3 == "-99", adm0_a3, iso_a3),
    name
  ) %>%
  dplyr::distinct() %>%
  dplyr::filter(!is.na(iso_a3), iso_a3 != "-99")


# =========================================================
# 7) Country-level latest cumulative deaths per million
#   - Filters to country rows (continent != NA)
#   - Selects latest date per ISO3 country
# =========================================================
deaths_country <- df %>%
  dplyr::filter(!is.na(continent)) %>%
  dplyr::filter(!is.na(iso_code)) %>%
  dplyr::group_by(iso_code) %>%
  dplyr::filter(date == suppressWarnings(max(date, na.rm = TRUE))) %>%
  dplyr::summarise(
    country          = first_non_na(location_clean),
    latest_date      = suppressWarnings(max(date, na.rm = TRUE)),
    deaths_pm_latest = first_non_na(total_deaths_per_million),
    .groups = "drop"
  ) %>%
  dplyr::filter(!is.na(deaths_pm_latest)) %>%
  dplyr::mutate(
    hover = paste0(
      "<b>", country, "</b>",
      "<br>ISO3: ", iso_code,
      "<br>Latest date: ", as.character(latest_date),
      "<br>Cumulative deaths per million: ",
      format(round(deaths_pm_latest, 1), nsmall = 1, big.mark = ",")
    )
  )

latest_date_overall <- suppressWarnings(max(df$date, na.rm = TRUE))


# =========================================================
# 8) Polygons missing from mapped data (diagnostic join)
# =========================================================
missing_on_map <- world_index %>%
  dplyr::left_join(
    deaths_country %>% dplyr::transmute(iso_a3 = iso_code, has_deaths = TRUE),
    by = "iso_a3"
  ) %>%
  dplyr::filter(is.na(has_deaths)) %>%
  dplyr::select(iso_a3, name) %>%
  dplyr::left_join(loc_diag_iso, by = c("iso_a3" = "iso_code_any")) %>%
  dplyr::mutate(
    reason = dplyr::case_when(
      !is.na(any_deaths_missing)        & any_deaths_missing        ~ "total_deaths_per_million missing in dataset for this country",
      !is.na(any_excluded_by_continent) & any_excluded_by_continent ~ "Present in dataset but excluded by continent==NA filter",
      !is.na(any_iso_failed)            & any_iso_failed            ~ "ISO conversion failed in dataset (name mismatch / non-standard label)",
      TRUE ~ "No matching deaths-per-million record in dataset"
    ),
    hover = paste0(
      "<b>", name, "</b>",
      "<br>ISO3: ", iso_a3,
      "<br>Status: Missing from deaths-per-million map",
      "<br>Reason: ", reason
    )
  ) %>%
  dplyr::select(iso_a3, name, reason, hover)


# =========================================================
# 9) Diagnostics exports (next to this script)
# =========================================================
readr::write_csv(
  deaths_country %>%
    dplyr::select(country, iso_code, latest_date, deaths_pm_latest) %>%
    dplyr::arrange(country),
  out_path("non_missing_cum_deaths_per_million_countries.csv")
)

readr::write_csv(
  missing_on_map %>%
    dplyr::rename(iso_code = iso_a3, country = name) %>%
    dplyr::arrange(country),
  out_path("missing_cum_deaths_per_million_or_join_countries.csv")
)

iso_failed_locations <- loc_diag %>%
  dplyr::filter(reason == "ISO conversion failed (countrycode could not map name to ISO3)") %>%
  dplyr::select(location, location_clean, reason) %>%
  dplyr::arrange(location_clean)

readr::write_csv(
  iso_failed_locations,
  out_path("iso_conversion_failed_locations.csv")
)

all_world_polygons_with_status <- world_index %>%
  dplyr::left_join(
    deaths_country %>% dplyr::select(iso_code, country, latest_date, deaths_pm_latest),
    by = c("iso_a3" = "iso_code")
  ) %>%
  dplyr::left_join(loc_diag_iso, by = c("iso_a3" = "iso_code_any")) %>%
  dplyr::mutate(
    status = dplyr::if_else(!is.na(deaths_pm_latest), "Non-missing (mapped)", "Missing (grey)"),
    reason = dplyr::case_when(
      status == "Non-missing (mapped)"                                  ~ "OK",
      !is.na(any_deaths_missing)        & any_deaths_missing            ~ "total_deaths_per_million missing in dataset for this country",
      !is.na(any_excluded_by_continent) & any_excluded_by_continent     ~ "Present in dataset but excluded by continent==NA filter",
      !is.na(any_iso_failed)            & any_iso_failed                ~ "ISO conversion failed in dataset (name mismatch / non-standard label)",
      TRUE                                                           ~ "Not present in dataset (or not convertible via dataset location names)"
    )
  ) %>%
  dplyr::transmute(
    iso_code      = iso_a3,
    polygon_name  = name,
    status,
    reason,
    latest_date,
    deaths_pm_latest
  ) %>%
  dplyr::arrange(status, polygon_name)

readr::write_csv(
  all_world_polygons_with_status,
  out_path("all_world_polygons_with_status.csv")
)

cat("Diagnostics exported to:\n")
cat(" - ", out_path("non_missing_cum_deaths_per_million_countries.csv"), "\n", sep = "")
cat(" - ", out_path("missing_cum_deaths_per_million_or_join_countries.csv"), "\n", sep = "")
cat(" - ", out_path("iso_conversion_failed_locations.csv"), "\n", sep = "")
cat(" - ", out_path("all_world_polygons_with_status.csv"), "\n\n", sep = "")

cat("Latest date overall in dataset: ", as.character(latest_date_overall), "\n", sep = "")
cat("ISO conversion failures (rows): ", sum(is.na(df$iso_code)), "\n", sep = "")
cat("Unique locations failing ISO conversion: ", nrow(iso_failed_locations), "\n\n", sep = "")


# =========================================================
# 10) Map data prep (NA retained for missing; no 0 imputation)
# =========================================================
map_data <- world_index %>%
  dplyr::left_join(
    deaths_country %>% dplyr::select(iso_code, deaths_pm_latest, hover),
    by = c("iso_a3" = "iso_code")
  ) %>%
  dplyr::mutate(
    hover = dplyr::if_else(
      is.na(deaths_pm_latest),
      paste0(
        "<b>", name, "</b>",
        "<br>ISO3: ", iso_a3,
        "<br>Status: Missing cumulative deaths per million"
      ),
      hover
    )
  )

map_non_missing <- map_data %>% dplyr::filter(!is.na(deaths_pm_latest))
map_missing     <- map_data %>% dplyr::filter(is.na(deaths_pm_latest))


# =========================================================
# 11) Palette + robust zmax (99th percentile)
# =========================================================
colorscale_deaths <- list(
  list(0.00, "#FFFFCC"),
  list(0.20, "#FFEDA0"),
  list(0.35, "#FED976"),
  list(0.50, "#FEB24C"),
  list(0.65, "#FD8D3C"),
  list(0.80, "#FC4E2A"),
  list(0.92, "#E31A1C"),
  list(1.00, "#800026")
)

zmax_deaths <- stats::quantile(map_non_missing$deaths_pm_latest, probs = 0.99, na.rm = TRUE)
if (!is.finite(zmax_deaths) || zmax_deaths <= 0) {
  zmax_deaths <- max(map_non_missing$deaths_pm_latest, na.rm = TRUE)
}


# =========================================================
# 12) Flat world map (2 traces: coloured + grey overlay)
# =========================================================
base_geo_flat <- list(
  showframe      = FALSE,
  showcoastlines = FALSE,
  projection     = list(type = "natural earth", scale = 1.15),
  showocean      = TRUE,
  oceancolor     = "#FAFAFA",
  showland       = TRUE,
  landcolor      = "#FFFFFF",
  showcountries  = TRUE,
  countrycolor   = "#FFFFFF"
)

map_flat <- plotly::plot_ly() %>%
  plotly::add_trace(
    data         = map_non_missing,
    type         = "choropleth",
    locations    = ~iso_a3,
    locationmode = "ISO-3",
    z            = ~deaths_pm_latest,
    text         = ~hover,
    hoverinfo    = "text",
    colorscale   = colorscale_deaths,
    autocolorscale = FALSE,
    zauto        = FALSE,
    zmin         = 0,
    zmax         = zmax_deaths,
    showscale    = TRUE,
    colorbar     = list(title = "Cumulative deaths / 1M", tickformat = ",.0f"),
    marker       = list(line = list(color = "#FFFFFF", width = 0.3))
  ) %>%
  plotly::add_trace(
    data         = map_missing,
    type         = "choropleth",
    locations    = ~iso_a3,
    locationmode = "ISO-3",
    z            = I(1),
    text         = ~hover,
    hoverinfo    = "text",
    colorscale   = list(list(0, "#BDBDBD"), list(1, "#BDBDBD")),
    showscale    = FALSE,
    marker       = list(line = list(color = "#FFFFFF", width = 0.3))
  ) %>%
  plotly::layout(
    title  = paste0(
      "Cumulative COVID-19 deaths per million",
      "<br>(flat world; overall latest = ",
      as.character(latest_date_overall),
      ")"
    ),
    geo    = base_geo_flat,
    autosize = FALSE,
    height = 500,
    margin = list(l = 0, r = 0, t = 50, b = 0)
  )

if (knitr::is_latex_output()) {
  if (!requireNamespace("webshot2", quietly = TRUE)) {
    stop("Package 'webshot2' is required to include Plotly widgets in PDF. Install with install.packages('webshot2').")
  }

  flat_tmp_html <- out_path("tmp_flat_map.html")
  flat_png <- out_path("CumDeathsPerMillion_FlatWorld.png")

  htmlwidgets::saveWidget(map_flat, flat_tmp_html, selfcontained = TRUE)
  webshot2::webshot(
    url      = flat_tmp_html,
    file     = flat_png,
    selector = "div.html-widget",
    zoom     = 2
  )

  knitr::include_graphics(flat_png)
} else {
  map_flat
}

flat_html   <- out_path("CumDeathsPerMillion_FlatWorld.html")
flat_libdir <- out_path("CumDeathsPerMillion_FlatWorld_files")

htmlwidgets::saveWidget(
  widget         = map_flat,
  file           = flat_html,
  selfcontained  = FALSE,
  libdir         = flat_libdir
)

# browseURL(flat_html)


# =========================================================
# 13) Rotatable globe (orthographic; same 2-trace logic)
# =========================================================
geo_globe <- list(
  projection = list(
    type = "orthographic",
    rotation = list(lon = 10, lat = 20),
    scale = 1.15
  ),
  showframe      = FALSE,
  showocean      = TRUE,
  oceancolor     = "#FAFAFA",
  showland       = TRUE,
  landcolor      = "#FFFFFF",
  showcoastlines = TRUE,
  coastlinecolor = "rgba(0,0,0,0.15)",
  showcountries  = TRUE,
  countrycolor   = "#FFFFFF",
  showlakes      = TRUE,
  lakecolor      = "#EBF5FF"
)

map_globe <- plotly::plot_ly() %>%
  plotly::add_trace(
    data         = map_non_missing,
    type         = "choropleth",
    locations    = ~iso_a3,
    locationmode = "ISO-3",
    z            = ~deaths_pm_latest,
    text         = ~hover,
    hoverinfo    = "text",
    colorscale   = colorscale_deaths,
    autocolorscale = FALSE,
    zauto        = FALSE,
    zmin         = 0,
    zmax         = zmax_deaths,
    showscale    = TRUE,
    colorbar     = list(title = "Cumulative deaths / 1M", tickformat = ",.0f"),
    marker       = list(line = list(color = "#FFFFFF", width = 0.4))
  ) %>%
  plotly::add_trace(
    data         = map_missing,
    type         = "choropleth",
    locations    = ~iso_a3,
    locationmode = "ISO-3",
    z            = I(1),
    text         = ~hover,
    hoverinfo    = "text",
    colorscale   = list(list(0, "#BDBDBD"), list(1, "#BDBDBD")),
    showscale    = FALSE,
    marker       = list(line = list(color = "#FFFFFF", width = 0.4))
  ) %>%
  plotly::layout(
    title  = paste0(
      "Cumulative COVID-19 deaths per million",
      "<br>(rotatable globe; overall latest = ",
      as.character(latest_date_overall),
      ")"
    ),
    geo    = geo_globe,
    autosize = FALSE,
    height = 500,
    margin = list(l = 0, r = 0, t = 50, b = 0)
  ) %>%
  plotly::config(displayModeBar = TRUE, scrollZoom = TRUE)

if (knitr::is_latex_output()) {
  if (!requireNamespace("webshot2", quietly = TRUE)) {
    stop("Package 'webshot2' is required to include Plotly widgets in PDF. Install with install.packages('webshot2').")
  }

  globe_tmp_html <- out_path("tmp_globe_map.html")
  globe_png <- out_path("CumDeathsPerMillion_RotatableGlobe.png")

  htmlwidgets::saveWidget(map_globe, globe_tmp_html, selfcontained = TRUE)
  webshot2::webshot(
    url      = globe_tmp_html,
    file     = globe_png,
    selector = "div.html-widget",
    zoom     = 2
  )

  knitr::include_graphics(globe_png)
} else {
  map_globe
}

globe_html   <- out_path("CumDeathsPerMillion_RotatableGlobe.html")
globe_libdir <- out_path("CumDeathsPerMillion_RotatableGlobe_files")

htmlwidgets::saveWidget(
  widget        = map_globe,
  file          = globe_html,
  selfcontained = FALSE,
  libdir        = globe_libdir
)

# browseURL(globe_html)
```

